Understanding Factors Contributing to Application Approval in Urban Planning Decisions (2016-2024)

### 2024-12-02 paper/paper.qmd
Upload accuracy result rds file

The remainder of this paper is structured as follows. @sec-data introduces the overview of the data (@sec-data-overview), measurement (@sec-data-measurement), data cleaning process (@sec-data-cleaning), data features (@sec-data-feature), as well as explanations, descriptions, table and graph summaries of outcome (@sec-data-outcome) and predictor variables (@sec-data-predictor) of the study. @sec-model explains the modeling process in detail, including procedure of model set-up, model justifications, model assumptions, and potential limitations. Then, @sec-result highlights the prediction outcome and results by model summary and plots. Finally, @sec-dis includes discussions based on the results, linking it to the real world and also states some strengths and weakness of the study.

@@

Applications to the Committee of Adjustment are typically submitted by property owners, developers, or their authorized agents (e.g., architects or planners). Eligibility to submit an application requires ownership or legal rights to the property in question. For some application types, such as consents, additional prerequisites include demonstrating compliance with existing zoning bylaws or providing documentation showing that the proposed subdivision aligns with city planning objectives. Minor variance applications, on the other hand, require applicants to show that the requested adjustments are “minor” in nature, do not conflict with broader zoning intentions, and maintain compatibility with the surrounding area. As a result, individuals submitting these applications are often professionals with expertise in urban planning, real estate, or development, or property owners working in consultation with such professionals.

The data collection process begins with property owners or developers submitting applications through a centralized system. Each application includes essential details, which are systematically verified and recorded by administrative staff. Following public hearings, committee decisions (approval or refusal) are added to the database, completing the data entry process. The dataset is then maintained and periodically published by the City of Toronto, ensuring it remains accessible and accurate for public and analytical use.

@@

To ensure the robustness and reliability of the Bayesian logistic regression model, a series of validation and diagnostic steps were conducted. Detailed plots and results, including posterior predictive checks (PPC), posterior vs. prior comparisons, and convergence diagnostics, are provided in the appendix for further examination. Below, the focus is on other key validation methods:

1. **Out-of-Sample Testing**

The dataset was split into a training set (80%) and a test set (20%) to evaluate the model's predictive performance on unseen data. The model was trained on the training set, and predictions were made on the test set. Classification accuracy was calculated as the proportion of correctly classified outcomes in the test set. This approach assesses the model's generalizability.

2. **Confusion Metrix**
The confusion matrix was computed to evaluate the model's classification performance, providing measures such as accuracy, sensitivity, and specificity. This validation assesses the model's predictive capabilities and its performance in correctly classifying approval decisions.

@@

2. **Methodological Constraints**
The Bayesian logistic regression model assumes that the relationships between predictors and the outcome variable are additive and linear on the log-odds scale. This assumption may oversimplify real-world interactions between variables, such as how specific planning districts may interact with application types or temporal trends. Furthermore, the model does not fully account for potential multicollinearity among predictors, which could distort the estimated coefficients. While priors were carefully chosen, their influence on the results might vary with a different prior specification, introducing some degree of subjectivity into the analysis.

@@

## Idealized Methodology and Survey

### Overview
This section outlines an idealized methodology and survey design to supplement the observational dataset used in this study. The objective is to enhance the analysis of urban planning decisions by incorporating a carefully designed survey targeting key stakeholders in Toronto's zoning and land-use processes.

### Sampling Approach

#### Target Population
The target population comprises stakeholders involved in the Committee of Adjustment's decision-making process, including:

@@

#### Sampling Frame
The sampling frame would be constructed using publicly available application records and contact information from municipal directories. This ensures representation from all planning districts and captures diverse perspectives across geographic and professional domains.

#### Sample Size
Using stratified random sampling, the sample size would aim for approximately 3000 respondents, ensuring statistical validity while considering feasibility.

#### Stratified Random Sampling
Stratified random sampling enhances the accuracy of survey estimates because it ensures proportional representation of different demographic groups. This minimizes the risk of sampling bias, where certain groups may be overrepresented or underrepresented, skewing the results.

@@

#### Trade-offs
While stratified random sampling ensures proportional representation and enhances the precision of survey estimates, it is not without challenges. One notable trade-off is the complexity involved in defining and managing multiple strata, particularly when the strata are based on multiple dimensions such as application type, planning district, and applicant role. This complexity increases the time and resources required for sample design and execution. For instance, ensuring proportional representation across strata like **Scarborough developers** or **North York consultants** may necessitate oversampling in smaller groups, leading to additional effort in recruitment and logistical coordination.

@@

### Data Validation
Survey responses would be validated through cross-referencing with the cleaned dataset. For instance:

@@

### Survey

@@

#### Survey Structure

@@

### Surevey Design Consideration

@@

### Strength

@@

### Weakness

>>>>

This study is important because it addresses an area of urban development that has direct implications for policy-making and planning efficiency. By analyzing approval patterns, the research helps clarify how procedural factors, geographic contexts, and temporal variations affect outcomes, providing a deeper understanding of the mechanisms at work in planning decisions. This understanding is essential for fostering accountability and ensuring that urban growth aligns with community needs and regulatory frameworks. The findings also serve as a foundation for improving processes and reducing inconsistencies, contributing to more equitable and transparent urban governance.

The remainder of this paper is structured as follows. @sec-data introduces the overview of the data (@sec-data-overview), measurement (@sec-data-measurement), data cleaning process (@sec-data-cleaning), data features (@sec-data-feature), as well as explanations, descriptions, table and graph summaries of outcome (@sec-data-outcome) and predictor variables (@sec-data-predictor) of the study. @sec-model explains the modeling process in detail, including procedure of model set-up, model justifications, model assumptions, and potential limitations. Then, @sec-result highlights the prediction outcome and results by model summary and plots. Finally, @sec-dis includes discussions based on the results, linking it to the real world and also states some weakness of the study.

@@

Applications to the Committee of Adjustment are typically submitted by property owners, developers, or their authorized agents (e.g. architects or planners). Eligibility to submit an application requires ownership or legal rights to the property in question. There are two types of application, minor variance (MV) and consents (CO). Consent applications address land subdivision for developments or sales, while Minor Variance applications request small adjustments to zoning regulations like setbacks or height limits. For consent applications, additional prerequisites include demonstrating compliance with existing zoning bylaws or providing documentation showing that the proposed subdivision aligns with city planning objectives. Minor variance applications, on the other hand, require applicants to show that the requested adjustments are “minor” in nature, do not conflict with broader zoning intentions, and maintain compatibility with the surrounding area. As a result, individuals submitting these applications are often professionals with expertise in urban planning, real estate, or development, or property owners working in consultation with such professionals.

Applications to the Committee of Adjustment can be submitted for properties located in one of Toronto’s defined planning districts: Etobicoke York, North York, Scarborough, and Toronto East York. Each district represents a distinct geographic area within the city, characterized by varying levels of urban density, development patterns, and zoning requirements. The choice of district is determined by the location of the property for which the application is being made.

For the submission year, the dataset captures applications over a range of years, reflecting historical patterns and trends in urban development. Applications from earlier years may correspond to a period of slower development or more stringent zoning regulations, while more recent submissions could align with economic growth, population increases, or shifts in city planning priorities. For instance, years coinciding with major municipal initiatives or economic booms may see spikes in application numbers. By including the year of submission, the dataset provides a temporal dimension that connects development trends to broader economic and policy contexts, such as changes in zoning laws or housing demand.

The data collection process begins with property owners or developers submitting applications through a centralized system. Application based surveys are collected. Each application includes essential details, which are systematically verified and recorded by administrative staff. Following public hearings, committee decisions (approval or refusal) are added to the database, completing the data entry process. The dataset is then maintained and periodically published by the City of Toronto, ensuring it remains accessible and accurate for public and analytical use.

@@

To ensure the robustness and reliability of the Bayesian logistic regression model, a series of validation and diagnostic steps were conducted. Detailed plots and results, including posterior predictive checks (PPC), posterior vs. prior comparisons, and convergence diagnostics, are provided in the appendix for further examination. Below, the focus is on key validation method of **out of sampling**:

The dataset was split into a training set (80%) and a test set (20%) to evaluate the model's predictive performance on unseen data. The model was trained on the training set, and predictions were made on the test set. Classification accuracy was calculated as the proportion of correctly classified outcomes in the train set and test set. This approach assesses the model's generalizability.

```{r}
#| echo: false
#| eval: true
#| label: tbl-accuracy
#| tbl-cap: "Classification Accuracy for Model Validation"
#| warning: false

# Load the saved results
accuracy_results <- readRDS(file = here::here("models/accuracy_results.rds"))

# Extract the values
accuracy_train <- accuracy_results$accuracy_train
accuracy_test <- accuracy_results$accuracy_test

#### Summary Statistics for Accuracy ####
accuracy_summary <- tibble(
  `Train Accuracy` = accuracy_train,
  `Test Accuracy` = accuracy_test
)

#### Create a kable table ####
accuracy_summary %>%
  kable(
    format = "html"
  )
```

@tbl-accuracy presents the train and test accuracy. The model's training accuracy of 66.75% and test accuracy of 66.575% are quite close, indicating that the model generalizes reasonably well between the training and test datasets. However, the relatively low accuracy on both datasets suggests the presence of underfitting, where the model fails to capture the complexity of the data and adequately explain the variance in the outcomes. This could result from the model being too simplistic, with insufficient predictors or inadequate representation of the data's structure. Additionally, potential issues like noisy or unbalanced data, insufficient feature engineering, or a lack of higher-order interactions and non-linearities in the model could contribute to this underfitting.

@@

2. **Methodological Constraints**
The Bayesian logistic regression model assumes that the relationships between predictors and the outcome variable are additive and linear on the log-odds scale. This assumption may oversimplify real-world interactions between variables, such as how specific planning districts may interact with application types or temporal trends. The model's relatively low accuracy for both the training and test datasets suggests underfitting, indicating that it fails to capture the full complexity of the data and may not adequately explain the variance in outcomes. This limitation could stem from insufficient predictors, unaccounted non-linear relationships, or interactions among variables. Furthermore, the model does not fully account for potential multicollinearity among predictors, which could distort the estimated coefficients. While priors were carefully chosen, their influence on the results might vary with a different prior specification, introducing some degree of subjectivity into the analysis.

@@

# Idealized Methodology and Survey

## Overview
This section outlines an idealized methodology and survey design to supplement the observational dataset used in this study. The objective is to enhance the analysis of urban planning decisions by incorporating a carefully designed survey targeting key stakeholders in Toronto's zoning and land-use processes.

## Sampling Approach

### Target Population
The target population comprises stakeholders involved in the Committee of Adjustment's decision-making process, including:

@@

### Sampling Frame
The sampling frame would be constructed using publicly available application records and contact information from municipal directories. This ensures representation from all planning districts and captures diverse perspectives across geographic and professional domains.

### Sample Size
Using stratified random sampling, the sample size would aim for approximately 3000 respondents, ensuring statistical validity while considering feasibility.

### Stratified Random Sampling
Stratified random sampling enhances the accuracy of survey estimates because it ensures proportional representation of different demographic groups. This minimizes the risk of sampling bias, where certain groups may be overrepresented or underrepresented, skewing the results.

@@

### Trade-offs
While stratified random sampling ensures proportional representation and enhances the precision of survey estimates, it is not without challenges. One notable trade-off is the complexity involved in defining and managing multiple strata, particularly when the strata are based on multiple dimensions such as application type, planning district, and applicant role. This complexity increases the time and resources required for sample design and execution. For instance, ensuring proportional representation across strata like **Scarborough developers** or **North York consultants** may necessitate oversampling in smaller groups, leading to additional effort in recruitment and logistical coordination.

@@

## Data Validation
Survey responses would be validated through cross-referencing with the cleaned dataset. For instance:

@@

## Survey

@@

### Survey Structure

@@

## Surevey Design Consideration

@@

## Strength

@@

## Weakness
